---
title: "Statistical tests as linear models - unrankling ranks"
output: html_notebook
---

```{r, message=FALSE, echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(broom)
library(MASS)
```
# The premise
Jonas Lindeløv wrote a [nice blog post](https://lindeloev.github.io/tests-as-linear/#3_pearson_and_spearman_correlation) explaining how common statistical tests amount to special cases of linear regression. This is real neat unfication of tests that at first pass look unrelated. His work included examples of nonparametric tests, based on ranks. But he seems to dismiss nonparametric tests as not that different from parametric ones. That seemed off. But I realized that my intuition for what ranks do to datawas too limited to say for sure. So I made some examples here to try and understand ranks better.


# Verification

First, I just want to try a couple examples of parametric tests from Jonas' post to make sure things check out.

## Correlation (Pearsons)
We'll need 20 samples for the equivilance to be exact.
```{r, message=FALSE, echo=TRUE, fig.width=0.8, fig.height=0.8}
# Create some data that covaries
covar <- matrix(c(1, 0.8, 1, 0.8), ncol=2)
M <- mvrnorm(n = 20, rep(0, 2), Sigma=covar)
X <- M[,1]
Y <- M[,2]


M %>% 
  as.data.frame() %>% 
  ggplot(aes(x=V1, y=V2)) +
  geom_point() + 
  theme_classic() +
  labs(x="X", y="Y")
```

```{r, message=FALSE, echo=TRUE}
# Built in Pearons
a <- cor.test(X,Y)
at <- tidy(a)

# Equivalent linear model
b <- lm(Y ~ 1 + X)
bt = tidy(b)[2, ]  # Only slope
bt$conf.low = confint(b)[2,1]
bt$conf.high = confint(b)[2,2]
bt$method <- "lm(y ~ 1 + x)"

print(full_join(at, bt))
```

## _t_-test (one variable)
We'll need 20 samples for the equivilance to be exact.

```{r, message=FALSE, echo=FALSE}
Y1 <- rnorm(50, mean=1, sd=1)

# Built-in t-test
a = t.test(Y1)
at <- tidy(a)
at$df = a$parameter

# Equivalent linear model
b = lm(Y1 ~ 1)
bt <- tidy(b)
bt$conf.low = confint(b)[1]
bt$conf.high = confint(b)[2]
bt$df = b$df.residual
bt$method <- "lm(y ~ 1)"

print(full_join(at, bt))
```

## Paired _t_-test (two variable)
We'll need 20 samples for the equivilance to be exact.

```{r, message=FALSE, echo=FALSE}
Y0 <- rnorm(20, mean=0, sd=1)
Y1 <- rnorm(20, mean=1, sd=1)

# Built-in t-test
a = t.test(Y1, Y0, paired = TRUE)
at <- tidy(a)
at$df = a$parameter

# Equivalent linear model
b = lm(Y1 - Y0 ~ 1)
bt <- tidy(b)
bt$conf.low = confint(b)[1]
bt$conf.high = confint(b)[2]
bt$df = b$df.residual
bt$method <- "lm(y1 - y0 ~ 1)"

print(full_join(at, bt))
```

Now let's try their nonparametric equivilants.

HERE


# Ranks 
What does ranking do to data? 

To find out let's first try some different noise distributions, then some outliers, and finally some linear and nonlinear series.

## Gaussian noise

## Non-gaussian noise

## OMG Outliers

## Linear data

## Nonlinear data