---
title: "Statistical tests as linear models"
output: html_notebook
---

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(broom)
library(MASS)
```
# The premise
Jonas Lindeløv wrote a nice blog post explaining how many common statistical tests can be seen as special cases in simple linear regression. This included examples of nonparametric tests based on ranks. I realized in reading this my intuition for what ranks do to data was too limited. So I made some examples to try and better understand.


# Verifcation

First, I just want to try a few examples of parametric tests from Jonas' post to make sure things check out.

## Pearsons's
We'll need 20 samples for the equivilance to be exact.
```{r, message=FALSE, echo=FALSE}
# Create some data that covaries
covar <- matrix(c(1, 0.8, 1, 0.8), ncol=2)
M <- mvrnorm(n = 20, rep(0, 2), Sigma=covar)
X <- M[,1]
Y <- M[,2]

# Built in Pearons
a <- cor.test(X,Y)
at <- tidy(a)

# Equivalent linear model
b <- lm(Y ~ 1 + X)
bt = tidy(b)[2, ]  # Only slope
bt$conf.low = confint(b)[2,1]
bt$conf.high = confint(b)[2,2]
bt$method <- "lm(y ~ 1 + x)"

print(full_join(at, bt))
```

## t-test
We'll need 20 samples for the equivilance to be exact.

```{r, message=FALSE, echo=FALSE}
Y1 <- rnorm(50, mean=1, sd=1)

# Built-in t-test
a = t.test(Y1)
at <- tidy(a)
at$df = a$parameter

# Equivalent linear model
b = lm(Y1 ~ 1)
bt <- tidy(b)
bt$conf.low = confint(b)[1]
bt$conf.high = confint(b)[2]
bt$df = b$df.residual
bt$method <- "lm(y ~ 1)"

print(full_join(at, bt))
```



# Ranks 
## Gaussian noise

## Non-gaussian noise

## OMG Outliers

## Linear data

## Nonlinear data